|

### 2024-05-28 Drafts/Paper_1/01-analysis_no_age_correction.Rmd
Full draft!

save_plot(paste0(loc,"/Figures/tuned/No_age_correction/Lambda_boxplot_other_species_",what.2.tune,".png"),new.lam.plt.6,base_width = 19,base_height = 15)

>>>>

save_plot(paste0(loc,"/Figures/tuned/No_age_correction/Lambda_boxplot_other_species_",what.2.tune,".png"),new.lam.plt.others,base_width = 19,base_height = 15)

### 2024-05-28 Drafts/Paper_1/02-intro.Rmd
Full draft!

The view that marine populations are highly resilient likely originates from the views of some of the first fisheries scientists who suggested that marine fisheries were effectively inexhaustible [e.g., mcintoshFisheriesInterNationalCouncil1919a; @mcintoshFisheriesInternationalCouncil1919]. These ideas were subsequently married to the theoretical results from some of the earliest population dynamic models [e.g., @verhulstNoticeLoiQue1838] and lead to the development of *modern* fisheries science and MSY theory [@rickerStockRecruitment1954; @schaeferAspectsDynamicsPopulations1954; @bevertonDynamicsExploitedFish1957; ]. Fundamentally, the belief was that the annual reproductive output and/or juvenile survival rates would increase as abundance declines, otherwise fisheries would collapse (REF), while the old paradigm that fisheries were inexhaustible was no longer supported, the new paradigm indicated that increased harvesting activity would result in the increased *productivity* of a population. The subsequent models developed assumed that marine fishes would retain a latent capacity for increase, even if managed in such a way that fisheries removals balance population growth (under equilibrium conditions). This means that populations always have the ability to increase in size, and their lifetime reproductive rates in the absence of fishing would be expected to be greater than one (ref, ref). At the time many of these fisheries scientists identified the challenges and limitations of their modelling frameworks, including concerns about how applicable these concepts were for species with different life history characteristics, e.g., how different would the fisheries impact be on long lived species with low fecundity and short lived species with high fecundity (CITE SOMETHING).

>>>>

The view that marine populations are highly resilient likely originates from the views of some of the first fisheries scientists who suggested that marine fisheries were effectively inexhaustible [e.g., mcintoshFisheriesInterNationalCouncil1919a; @mcintoshFisheriesInternationalCouncil1919]. These ideas were subsequently married to the theoretical results from some of the earliest population dynamic models [e.g., @verhulstNoticeLoiQue1838] and led to the development of *modern* fisheries science and MSY theory [@rickerStockRecruitment1954; @schaeferAspectsDynamicsPopulations1954; @bevertonDynamicsExploitedFish1957; ]. Fundamentally, the belief was that the annual reproductive output and/or juvenile survival rates would increase as abundance declines, otherwise fisheries would collapse (REF), while the old paradigm that fisheries were inexhaustible was no longer supported, the new paradigm indicated that increased harvesting activity would result in the increased *productivity* of a population. The subsequent models developed assumed that marine fishes would retain a latent capacity for increase, even if managed in such a way that fisheries removals balance population growth (under equilibrium conditions). This means that populations always have the ability to increase in size, and their lifetime reproductive rates in the absence of fishing would be expected to be greater than one (ref, ref). At the time, many of these fisheries scientists identified a number of challenges and limitations of their modelling frameworks, these concerns often anticipated many of the criticisms of fisheries science over the last 75 years.  These early concerns included questions regarding the applicability of these concepts for species with different life history characteristics, e.g., how different would the fisheries impact be on long lived species with low fecundity and short lived species with high fecundity (CITE SOMETHING).

### 2024-05-28 Drafts/Paper_1/05-discussion.Rmd
Full draft!

>>>>

Here we deconstructed the ecological demographics of numerous harvested marine fish populations . In general, these populations have latent capacity for growth and, even when accounting for the influence of fishing, these growth rates are very close to 1 in general.  There is evidence that fisheries stocks are fished in such a way that they will tend to decline until they have declined below 30% of the maximum population size observed, after that most fished populations tend to experience population growth rates above 1. Most populations appear to exhibit evidence for negative density dependence, which is in line with general beliefs within fisheries science and population ecology more generally, however there are some populations where the evidence suggests some level of density independence or weak positive density dependence , including most of the *G. morhua* populations in NW Atlantic.  Both variability and median $\lambda_{F=0}$ influence recovery, stocks with a relatively high variability in $\lambda_{F=0}$ had a moderate-high recovery potential even when their median $\lambda_{F=0}$ was below one. Effectively recovery of the stocks was likely within 20 years whenever the median $\lambda_{F=0}$ was greater than 1.06 or the standard deviation of lambda was greater than 0.2.

For every population in this analysis the $\lambda_{F=0}$ at the minimum observed abundance was estimated to be greater than one; suggesting all of the populations have the potential for some degree of recovery given enough time when there is no fishing mortality. However, there are seven populations in which $\lambda_{F=0}$ at 40% of the maximum abundance is below 1, five of these populations are found in the NW Atlantic, and five of these are in the Gadus genus (*G. morhua* (3) and *G. chalcogrammus* (2)). For both of these Gadus species, there are examples of populations in which the latent capacity for growth is far more robust.  For example, *G. chalcogrammus* is one of the largest fisheries in the world and the vast majority of the landings are from the Eastern Bering Sea population which has a high latent capacity for growth. However the $\lambda_{F=0}$ estimates for two neighboring populations in Alaska are much lower, and it is likely that these populations experience much lower fishing pressure because of their low latent capacity for growth.  
In contrast, in the NW Atlantic, several *G. morhua* populations supported large distant waters fishery for centuries (citations), however the data used in this analysis indicated the majority of the stocks in the NW Atlantic have minimal recovery potential. The maximum abundance used for these populations was likely a small fraction of the historic maximum abundance of these stocks, primarily due to over fishing by the industrial fishery that developed in the latter half of the 20th century (Cite Rose book maybe?, others). Here it is likely that the fishery induced changes to the population, which combined with changes in the environment (Cite cold water and seals) has resulted in these populations having little latent capacity for growth or recovery. Indeed, it has been suggested that the SGSL cod population will go extinct unless natural mortality declines significantly (Doug’s predator pit stuff); the results of this analysis largely align with these suggestions given the low recovery potential of this stock and it's extremely low latent capacity for growth. However, as one moves towards the eastern Atlantic, the low recovery potential of *G. morhua* populations was no longer observed, with stocks around mainland Europe all having a high recovery potential. As with *G. chalcogrammus*, this indicates that regional variability is a factor in the ability of populations to withstand fishing pressure. While some Gadus populations have an ability to withstand large fisheries due to their latent capacity for growth, the life-history at a species level is insufficient to predict how a population will respond to fishing. Moreover, this analysis is relevant for only the period under study, the recovery potential assumes that the underlying population demographic rates in the future will be similar to those observed in the past. However, as is likely the case with the *G. morhua* stocks in the NW Atlantic, the response for a population may change over time due to environmental, ecological, and anthropocentric factors (CITE!).

More generally, we observed that populations with high variability in $\lambda_{F=0}$ have a high recovery potential, however, this variability does not necessarily follow on from the expectations of conventional life-history theory (Cite Kingsvader). For example, *C. harengus* are a small, relatively short lived, low trophic level pelagic species (Cite!), based on life-history theory *C. harengus* should exhibit highly variable population growth potential in which they ‘exploit’ opportunities within the environment (CITE). However, the variability of $\lambda_{F=0}$ for the *C. harengus* populations was much lower than the *G. morhua* and *M. aeglefinus* populations found throughout the NE Atlantic. Other small pelagic populations, such at *A. tobianus* and *T. esmarkii* do have elevated  $\lambda_{F=0}$ as life-history theory would expect for small low trophic level species. However, caution would be advised in suggesting that other populations of these species would have similar population dynamics, especially given these two populations are found in the North Sea, which appears to be an environment in which high variability in $\lambda_{F=0}$ is commonplace. In addition, while stocks with a high variability in $\lambda_{F=0}$ do have years with very high population growth, these events happen at irregular intervals, even when populations are at low abundances. As such, for populations in which recovery plans and conservation targets are being developed, the developers must account for *both* the magnitude and the frequency of these events. Populations do not *always* increase at their maximum intrinsic rate of growth when at low abundance, but they may have the ability to do so occasionally.

The optimization procedure used for the life-table reconstruction herein assumes that both the estimates of abundance and the fishery removals from the stock assessment are correct. Given the underlying assessment models estimate biomass and abundance from fishery independent surveys that have undergone multiple years of peer review, we are confident that the underlying trends captured by the assessments provide a reasonable snapshot of the population dynamics of these population. However, it is likely that the fisheries removals used in the stock assessments do not capture all the mortality due to fishing, unreported fishing, bycatch and incidental mortality could all result in higher fishery removals than estimated in the assessments (CITE). This would result in the estimates of  $\lambda_{F=0}$ being lower than reality; ‘missing’ mortality would have been accounted for by reductions in fecundity and/or an increase in natural mortality. The optimization routine resulted in a median decrease in the no-fishing $\lambda_{F=0}$ estimates of approximately 6.8% when compared to the non-optimized estimates; the directionality of this change is consistent with fishing mortality being underestimated. Thus, $\lambda_{F=0}$ is likely an underestimate of the latent capacity for growth of these populations, this magnitude of this bias in any year would be dependent on the level of unaccounted for fishing mortality a population experienced.

Here we provide an overview of the population growth and recovery potential of numerous harvested marine fish populations. There is a latent capacity for growth for most populations and the majority of the populations appear to have the ability to recover relatively quickly if fishing mortality is reduced to low levels. However, there can be significant intra-species variability which may be related to environmental, ecological, and anthropocentric factors. This makes it difficult to use species level life-history traits to predict the population dynamics of specific populations.  The life-table reconstruction enables the underlying demographic rates to be estimated for populations with sufficient age-structured data, this provides an opportunity of understanding the dynamics of these populations using concepts developed in conservation biology and ecology.

## Ignore this, unless something is in here you want to pull into discussion
